---
title: "Raster-functions"
author: "Michiel P. Veldhuis"
date: "21 juni 2016"
output: html_document
---

Start by putting up objectives/tasks that students will be working though:
1. Import a raster — A lidar canopy height model (lidar/Teak_lidarCHM.tif)
1. For the CHM, set values == 0 to NA (not trees)
1. Classify the raster according to some distribution – low medium and tall trees. This could be done using a histogram potentially or we could just decide that 6m is generally grasses / understory, 10m small trees,and the rest are tall trees. A function could import the desired thresholds. 
1. Plot the classified raster, add a legend for each “class” - legends are super tricky to simplifying this process with a function would be good.  see: http://neon-workwithdata.github.io/neon-data-institute-2016/R/classify-by-threshold-R/  for my take on forcing a legend outside of the plot area using par settings. You may have other better forms of magic to make this work well. :)
1. Export the plot figure to a pdf – publishable
1. Export the classified raster as a geotiff with NaFlagg = -9999 to an outputs folder.

## Load libraries

```{r libraries}

# load libraries
library(raster)
library(rgdal)
source("scripts/functions.R")

```

## Import a raster — A lidar canopy height model (lidar/Teak_lidarCHM.tif)

```{r import-raster}

# Load raster
CHM_teak <- raster("../NEONdata/D17-California/TEAK/2013/lidar/TEAK_lidarCHM.tif")
chosen.breaks <- c(6,30,50,100)
# quick sanity check
plot(CHM_teak)

```

## For the CHM, set values == 0 to NA (not trees)

```{r Set 0 values to NA}

# Set 0 values to NA
CHM_teak[CHM_teak==0] <- NA
hist(CHM_teak)
plot(CHM_teak)

```

# Check density

```{r check-density}

density(CHM_teak)

plot_chm_density <- function(rast.in, title, bins) {
  density(rast.in, main=title, xlab = "Height (m)")
  abline(v = bins, col="red")
}

plot_chm_density(rast.in = CHM_teak, 
                 title = "Canopy heights at Teakettle", 
                 bins= chosen.breaks)


```

## Classify the raster according to some distribution – low medium and tall trees. This could be done using a histogram potentially or we could just decide that 6m is generally grasses / understory, 10m small trees,and the rest are tall trees. A function could import the desired thresholds. 

```{r reclassify-raster}

class.recl <- c(0, 6, 1,
                6, 30, 2,
                30, 50, 3,
                50, 70, 4)
class.recl
recl.matr <- matrix(class.recl, ncol=3, byrow=TRUE)
recl.matr

CHM_teak.recl <- reclassify(CHM_teak, recl.matr)
plot(CHM_teak.recl)

density(CHM_teak, 
        main = "Canopy heights at Teakettle site\n with bin cutoffs in red",
        xlab = "Height (m)")
abline(v = recl.matr[,2], col = "red")

```

## Save to pdf


```{r function-to-save-pdf}

make_pdf <- function(expr, filename, ..., verbose = TRUE) {
    if (verbose) {
        message("Creating: ", filename)
    }
    pdf(file = filename, ...)
    on.exit(dev.off())
    eval.parent(substitute(expr))}

make_pdf(plot_chm_density(rast.in = chm,
                          title = "Canopy heights at Teakettle",
                          bins = chosen.breaks),
         filename = "TEAK_CHM_density_with_breaks.pdf",
         width = 6, height = 7)

```

```{r save-pdf}

pdf(file = "Teak_CHM_density_with_breaks.pdf", width = 6, height = 7)
density(CHM_teak, 
        main = "Canopy heights at Teakettle site\n with bin cutoffs in red",
        xlab = "Height (m)")
abline(v = recl.matr[,2], col = "red")
dev.off()
```

## Plot with nice legend

```{r plot-with-legend}

par(xpd = FALSE, mar = c(5.1, 4.1, 4.1, 4.5))

plot(CHM_teak.recl,
     col = c ("yellow", "orange", "red", "blue"), # hard coding colors
     main = "Canopy height classes\n Lower Tekettle",
     legend = FALSE)
     

# enable plotting legend outside bounds
par(xpd = TRUE) 

# legend position
leg.x <- par()$usr[2] + 20
leg.y <- par()$usr[4] + 50 - (abs(par()$usr[3] - par()$usr[4]) / 2)

# add the legend
legend(leg.x, leg.y,
       legend = c("0-6 m", "6-30 m", "30-50 m", "50-100 m"),
       fill = c("yellow", "orange", "red", "blue"),
       bty = "n")

```

## Save a pdf

```{r}

pdf(file = "Canopy_height_map_reclassifief_TEAK.pdf", width = 7, height = 6)
par(xpd = FALSE, mar = c(5.1, 4.1, 4.1, 4.5))

plot(CHM_teak.recl,
     col = c ("yellow", "orange", "red", "blue"), # hard coding colors
     main = "Canopy height classes\n Lower Tekettle",
     legend = FALSE)
plot(extent(CHM_teak.recl), add=T, lwd=3)


# enable plotting legend outside bounds
par(xpd = TRUE) 

# legend position
leg.x <- par()$usr[2] + 20
leg.y <- par()$usr[4] + 50 - (abs(par()$usr[3] - par()$usr[4]) / 2)

# add the legend
legend(leg.x, leg.y,
       legend = c("0-6 m", "6-30 m", "30-50 m", "50-100 m"),
       fill = c("yellow", "orange", "red", "blue"),
       bty = "n")
dev.off()

```

## Save the reclassified raster to a GeoTIFF

```{r save-geotiff}

writeRaster(CHM_teak.recl,
            filename = "reclassified_CHM_TEAK_GeoTIFF.tif",
            format = "GTiff",
            options = "COMPRESS=LZW",
            overwrite = TRUE,
            NAflag = -9999)

```

```{r create-file-with-sessionInfo}

sink(paste0(format(Sys.time(), "%Y-%m-%d_%H%M%S"),
            "_sessionInfo.txt"))
sessionInfo()
sink()

```



```{r}

# function to create a reclassify matrix from a set of breaks
create_height_class_matrix <- function(chosen.breaks) { 
  # Get the length of the breaks vector to
  # figure how many classes we will have
  br.len <- length(chosen.breaks)
  
  # initialize heigth class vector with 0
  height.class.m <- c(0)
  
  # for input of breaks = c(6,30,50,100)
  # we would like to make something like this
  
  # c(0,6,1,
  #   6,30,2,
  #   30,50,3
  #   50,100,4)
  
  for (i in 1:br.len) {
    height.class.m <- c(height.class.m, chosen.breaks[i - 1], chosen.breaks[i], i)
  }
  
  reclass.height.mat <- matrix(height.class.m,
                               ncol =3,
                               byrow = TRUE)
  
  reclass.height.mat
  
}

create_height_class_matrix(chosen.breaks)

```







##########################################
##########################################
# function to plot the reclassified raster

plot_reclassified_raster <- function(rast.in, site.name, breaks){
        # this is a tricky bit because we need to out the legend
        # outside of the plot region
        
        # Get colors for plotting
        bin.colors <- rev(terrain.colors(length(breaks)))
        
        # make room for a legend
        
        par(xpd = FALSE, mar = c(5.1, 4.1, 4.1, 4.5))
        
        # plot
        plot(rast.in,
                 col = bin.colors,
                 main = paste("Canopy height classes \n", site.name),
                 legend = FALSE)
        
        # allow legend to plot outside of bounds
        par(xpd = TRUE)
        
        # legend x
        leg.x <- par()$usr[2] + 20
        
        # legend y
        leg.y <- par()$usr[4] + 50 - (abs(par()$usr[3] - par()$usr[4]) / 2) 
        
        # create legend text
        height.mat <- create_height_class_matrix(breaks)
        
        # initialize legend text
        legend.text <- c()
        
        for (i in 1:length(breaks)) {
                
                legend.text <- c(legend.text, 
                                                 paste0(height.mat[i, 1], "-", 
                                                            height.mat[i, 2], " m"))
        }
        
        # create the legend
        legend(leg.x, leg.y,  # set x,y legend location
                   legend = legend.text,  # make sure the order matches colors
                   fill = bin.colors,
                   bty = "n") # turn off border
        
        # turn off plotting outside of bounds
        par(xpd = FALSE)
}

